GEE models

### 2017-07-01 doc/manuscript.Rmd
homa-ir to homa2

Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time points over
6 years. Outcome variables were calculated using the OGTT values. IS was
assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function was
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, alanine
aminotransferase, and family history of diabetes. NEFA were analyzed as both
concentrations (nmol/mL) and proportions (mol%) of the total fraction.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

The primary outcome variables for this analysis were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.

>>>>

Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time points over
6 years. Outcome variables were calculated using the OGTT values. IS was
assessed using HOMA2-%S and the Matsuda index, while beta-cell function was
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, alanine
aminotransferase, and family history of diabetes. NEFA were analyzed as both
concentrations (nmol/mL) and proportions (mol%) of the total fraction.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA2-%S largely reflects hepatic insulin sensitivity, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

The primary outcome variables for this analysis were 
HOMA2-%S, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(HOMA2-%S) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.
